---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sharkyIBM

<!-- badges: start -->
<!-- badges: end -->

sharkyIBM includes code to simulate and sample a population of sharks, with individual "simsharks" growing, breeding and dying based on age- and/or sex-specific probability distributions governing growth, movement, fecundity, and mortality. The simulations track first and second-order relatives, and allow combinations of gene flow and sampling design that resemble a real population of interest e.g., via simulating multiple genetically distinct subpopulations that are sampled during an aggregation.

The primary purposes of this package are to 
1) assist with close-kin mark-recapture (CKMR) sample design, and 
2) develop/hone CKMR models.

## Installation

You can install the development version of sharkyIBM from [GitHub](https://github.com/) with:

``` r
# install.packages("pdevtools")
devtools::install_github("JDSwenson/sharkyIBM")
```

## Usage

Explain a potential use case ... 

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(sharkyIBM)
## basic example code
```

You'll need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this.

You can also embed plots, for example:

```{r pressure, echo = FALSE}
plot(pressure)
```
