---
title: "Using sharkyIBM: A Guide to Simulating and Sampling Dolphin Populations"
subtitle: "How to set up, calibrate, and run population simulations"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    fig-width: 9
    fig-height: 5
    code-fold: false
editor: 
  markdown: 
    wrap: 72
---

## Overview

**sharkyIBM** is an individual-based population simulation (IBM) package
designed to model dolphin populations with realistic life-history
dynamics and sampling schemes. The package implements a three-step
workflow:

1.  **Calibrate** a stable population using `create.stable.pop()`
2.  **Simulate** the population forward in time using `simulate.pop()`
3.  **Sample** from the population using `sample.pop()`

This guide walks you through each step using biologically realistic
parameters for **eastern spinner dolphins** (*Stenella attenuata*).

## Loading the Package

```{r}
#| label: setup
library(sharkyIBM)
library(tidyverse)
library(ggplot2)

# Set a seed for reproducibility
set.seed(7729)
```

## Part 1: Understanding the Population Model

### What is an Individual-Based Model (IBM)?

An IBM is a computational simulation where each animal is tracked
individually with its own characteristics (age, sex, reproductive state,
parentage). This allows us to capture individual variation and realistic
breeding patterns that would be difficult to model with population-level
equations.

In sharkyIBM: - Each dolphin has a unique ID, age, sex, maturity status,
and reproductive history - Dolphins age each year and die based on
age-specific survival probabilities - Females cycle through reproductive
states (pregnant, nursing calf, resting) - Offspring inherit their
mother's identity, creating a complete pedigree - The population can be
sampled in ways that mimic real fishing operations

### The Markov Breeding Cycle (Simplified Explanation)

The dolphin's reproductive life cycles through three
repeating states:

1.  **PREGNANT (S1):** Carrying a fetus for \~11 months. Once pregnant,
    she will give birth.
2.  **WITH DEPENDENT CALF (S2):** Just gave birth and nursing a calf for
    19-24 months. She is **suppressed** from breeding again —
    lactational suppression.
3.  **RESTING (S3):** Her calf has been weaned or died. She's ready to
    breed again.

The **key biological realism**: a mother whose calf dies is "released"
from lactational suppression and can breed sooner. A model that doesn't
capture this would miss an important ecological response to high calf
mortality.

Each year, the female's state is updated randomly based on her current
state and age-specific probabilities. The result: a distribution of
calving intervals emerges naturally from the model without being forced
in.

### Density Dependence (Simplified Explanation)

Imagine a population of dolphins. When populations are small (and
resources plentiful), dolphins can breed frequently. When populations
grow large (and food or space becomes scarce), breeding slows — fewer
conceptions occur, calving intervals get longer.

We model this using a mechanism called **Pella-Tomlinson compensation**.
As the population declines from its carrying capacity (K), conception
rates increase — the population "compensates" by "breeding harder" to
recover. This is done by adjusting the conception probabilities based on
population size (depletion).

**In plain terms:** When dolphins are relatively scarce, each female breeds more
often to help the population recover. When they're relatively abundant, breeding
slows. This is a realistic feedback loop that keeps the population
self-correcting.

## Part 2: Life-History Parameters for Eastern Spinner Dolphins

Based on the eastern spinner dolphin life-history model, here are the
recommended parameters:

### Mortality (Siler U-Shaped Curve)

Eastern spinners follow a classic mammalian mortality curve: - **High
calf mortality** (\~0.15/year excess) — calves are vulnerable - **Low
adult mortality** (\~0.04/year baseline) — prime-age dolphins live
well - **Senescent mortality** (\~0.10/year added) — old dolphins have
higher mortality

```{r}
#| label: mortality-params

# Age sequence
ages <- 0:40

# Siler parameters (U-shaped mortality)
M_adult <- 0.04       # Baseline adult mortality rate
M_calf_excess <- 0.15 # Additional mortality for calves
b_juv <- 0.5          # Juvenile decay rate (how fast juvenile mortality declines)
a_sen_onset <- 28     # Age when senescence begins
M_sen_add <- 0.10     # Mortality ADDED by senescence, attained at a_sen_ref (not at max_age)
sen_curve <- 0.15     # Curvature of the senescent rise; smaller = gentler upswing
a_sen_ref <- 40        # Reference age at which M_sen_add is fully realized (> a_sen_onset)

# Construct the full mortality curve. The senescent term is normalized so it
# is 0 at a_sen_onset and exactly M_sen_add at a_sen_ref, then keeps rising
# beyond it. sen_curve controls how quickly it climbs between those two anchors.
M_at_age <- M_calf_excess * exp(-b_juv * ages) + M_adult
sen_denom <- exp(sen_curve * (a_sen_ref - a_sen_onset)) - 1
M_at_age[ages >= a_sen_onset] <- M_at_age[ages >= a_sen_onset] +
  M_sen_add * (exp(sen_curve * (ages[ages >= a_sen_onset] - a_sen_onset)) - 1) /
              sen_denom

# Convert mortality to survival
survival <- exp(-M_at_age)

# Plot the survival curve
data.frame(age = ages, survival = survival) |>
  ggplot(aes(age, survival)) +
  geom_line(linewidth = 1) +
  labs(x = "Age (years)", y = "Annual Survival Probability",
       title = "Eastern Spinner Dolphin Survival Curve (Siler)",
       subtitle = "High juvenile mortality, flat in prime ages, senescence after age 28") +
  theme_minimal() +
  scale_y_continuous(limits = c(0, 1))
```

### Maturity (Logistic Ogive)

Eastern spinner females reach sexual maturity around age 9, with the
probability following a logistic (S-shaped) curve:

```{r}
#| label: maturity-params

# Logistic maturity ogive
a50 <- 9       # Age at 50% maturity
slope <- 1.5   # Steepness of the curve

ogive_f <- plogis(ages, location = a50, scale = 1/slope)

# Plot the maturity curve
data.frame(age = ages, maturity_prob = ogive_f) |>
  ggplot(aes(age, maturity_prob)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = a50, linetype = "dashed", color = "gray50") +
  labs(x = "Age (years)", y = "Probability of Being Mature",
       title = "Female Maturity Ogive (Logistic)",
       subtitle = "50% mature by age 9") +
  theme_minimal() +
  scale_y_continuous(limits = c(0, 1))
```

### Breeding Parameters

```{r}
#| label: breeding-params

# Lactational suppression (log-odds from observed pregnant:lactating ratio)
rho <- -3.255

# Calving interval and suppression ratio
target_interval <- 2.84      # Years between successive calves (observed)
target_depletion <- 0.3      # Depletion at which this interval was observed
suppression_ratio <- 0.15    # psi_nurse / psi_rest

# Pella-Tomlinson shape for density dependence
z_pt <- 2.39  # IWC convention; sets MNPL at 0.6 K

cat("Breeding parameters:\n")
cat(sprintf("  Lactational suppression (rho): %.3f\n", rho))
cat(sprintf("  Target calving interval: %.2f years\n", target_interval))
cat(sprintf("  Target depletion: %.1f%%\n", target_depletion * 100))
cat(sprintf("  Pella-Tomlinson shape (z): %.2f\n", z_pt))
```

### Demographic Parameters

```{r}
#| label: demo-params

max_age <- 40              # Maximum lifespan
litter_size <- 1           # Always one calf per pregnancy
female_fraction <- 0.5     # Sex ratio at birth
weaning_age <- 2L          # Years a calf is dependent (age 0-1)
num_mates <- 1L            # Polyandry level (1 = one sire per female)
infertility <- c(0.05, 0.05)  # Female, male infertility rates
```

## Part 3: Step-by-Step Workflow

### Step 1: Calibrate the Population

The first step calibrates a stable population using the
`create.stable.pop()` function. This function:

1.  **Phase 1 (Leslie):** Uses a Leslie matrix to estimate the age-0
    survival (s0) or density-dependence parameter (theta) analytically
2.  **Phase 2 (IBM):** Runs a simulation to refine the estimate until
    the population growth rate is essentially zero
3.  **Returns:** A configuration object bundling all parameters together

```{r}
#| label: calibrate-population
#| message: true

cat("Calibrating population...\n\n")

config <- create.stable.pop(
  # Life history
  max_age         = max_age,
  survival        = survival,
  maturity_age    = ogive_f,
  litter_size     = litter_size,
  
  # Breeding cycle
  rho             = rho,
  weaning_age     = weaning_age,
  
  # Demographics
  female_fraction = female_fraction,
  num_mates       = num_mates,
  infertility     = infertility,
  
  # Social structure
  pod_size        = 20,
  superpod_size   = 10,
  stickiness_year = 0.9,
  male_behavior   = "random",
  
  # Population size for calibration
  pop_size        = 20000,
  
  # Density dependence
  density_dependence = TRUE,
  z_pt            = z_pt,
  target_interval = target_interval,
  target_depletion = target_depletion,
  
  # Convergence
  check_interval  = 10L,
  growth_tol      = 0.001,
  stable_required = 5L,
  max_windows     = 100L
)

cat("\nCalibration complete!\n")
cat("Key outputs:\n")
cat(sprintf("  Calving interval at K: %.2f years\n", config$interval_K))
cat(sprintf("  dd_max (compensation strength): %.4f\n", config$dd_max))
cat(sprintf("  K_1plus (1+ animals at K): %s\n", format(config$K_1plus, big.mark = ",")))
```

### Step 2: Solve for Sustainable Fishing Mortality

If you want to fish the population at a stable level, use
`solve_F_sustainable()`:

```{r}
#| label: solve-F
#| message: true

cat("Solving for sustainable fishing mortality at D=0.3...\n\n")

f_result <- solve_F_sustainable(config)

cat("\nFishing solution:\n")
cat(sprintf("  F (Leslie analytical): %.4f\n", f_result$F_leslie))
cat(sprintf("  F (IBM-calibrated): %.4f (%.1f%% lower)\n", 
            f_result$F_sustainable,
            100 * (1 - f_result$F_sustainable / f_result$F_leslie)))
cat(sprintf("  Calving interval at equilibrium: %.2f years\n", 
            f_result$interval_at_equilibrium))
```

### Step 3: Run the Simulation

Simulate the population forward in time:

```{r}
#| label: run-simulation
#| message: true

cat("Simulating population for 100 years...\n\n")

sim <- simulate.pop(
  sim_config = config,
  num_years = 100,
  sample_years = c(50, 100),
  
  # Fishing parameters (optional)
  F_t = f_result$F_sustainable,
  selectivity = f_result$selectivity,
  init_depletion = 0.3
)

cat("\nSimulation complete!\n")

# Extract population summary
pop_summary <- sim$pop_summary

# Post-burn-in only
burn_in <- 2 * config$max_age
post_burn <- pop_summary[pop_summary$year > burn_in, ]

cat(sprintf("\nPost-burn-in population (years %d-%d):\n", 
            burn_in + 1, max(post_burn$year)))
cat(sprintf("  Initial N: %s\n", format(post_burn$N[1], big.mark = ",", trim = TRUE)))
cat(sprintf("  Final N: %s\n", format(tail(post_burn$N, 1), big.mark = ",", trim = TRUE)))
cat(sprintf("  Mean N: %s\n", format(round(mean(post_burn$N)), big.mark = ",", trim = TRUE)))

# Analyze stability
log_N <- log(post_burn$N)
growth_rates <- diff(log_N)
cat(sprintf("  Mean growth rate: %+.5f /year\n", mean(growth_rates)))
cat(sprintf("  SD growth rate: %.5f\n", sd(growth_rates)))
```

### Step 4: Visualize the Trajectory

```{r}
#| label: plot-trajectory
#| fig-cap: "Population trajectory during and after burn-in."

# Full population summary
pop_summary_all <- sim$pop_summary

# Compute target N at depletion
target_N <- config$pop_size * target_depletion

ggplot(pop_summary_all, aes(x = year, y = N)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = burn_in, linetype = "dotted", color = "gray50", linewidth = 0.5) +
  geom_hline(yintercept = target_N, linetype = "dashed", color = "red", alpha = 0.6) +
  labs(x = "Year", y = "Population Size (age 1+)",
       title = "Population Trajectory with Sustainable Fishing") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

### Step 5: Sample from the Population

Finally, sample individuals to simulate a fishing operation or genetic
sampling study:

```{r}
#| label: sample-population
#| message: true

cat("Sampling from population at year 100...\n\n")

samples <- sample.pop(
  sim_output = sim,
  n_trips = 5,
  n_sets = 4,
  sample_size = 30,
  sample_per = "set",
  sampling = "superpod",
  stickiness_set = 0.8,
  stickiness_trip = 0.6
)

cat(sprintf("\nSampling complete!\n"))
cat(sprintf("  Total sampled individuals: %d\n", nrow(samples)))
cat(sprintf("  Unique sampled individuals: %d\n", length(unique(samples$id))))
cat(sprintf("  Duplicates due to stickiness: %d\n", 
            nrow(samples) - length(unique(samples$id))))
```

## Key Parameters Summary

| Parameter | Value | Meaning |
|----|----|----|
| `max_age` | 40 | Maximum lifespan (years) |
| `survival` | Siler curve | Age-specific survival probabilities |
| `ogive_f` | Logistic (a50=9) | Female maturity curve |
| `litter_size` | 1 | Calves per pregnancy |
| `rho` | -3.255 | Log-odds of nursing vs resting conception |
| `target_interval` | 2.84 | Calving interval at target depletion |
| `target_depletion` | 0.3 | Reference depletion level |
| `z_pt` | 2.39 | Pella-Tomlinson shape |
| `weaning_age` | 2 | Years of calf dependency |
| `pod_size` | 20 | Family group size |
| `superpod_size` | 10 | Community size |

------------------------------------------------------------------------

**For more information:** See `?create.stable.pop()`, `?simulate.pop()`,
`?sample.pop()`
